Experiment 4 of 4

### 2016-09-27 experiments/e4/writeup/preprocessing/preprocessing.Rmd
less verbose e4 writeup

---
title: "Experiment 3 of 4"
output:
  html_document:
    toc: true
    <!--toc_float: true-->
    mathjax: default
    template: default
    code_folding: hide
    theme: default
---  
# Experiment 3

```{r docOpts}
library(knitr)
```

```{r loadLibraries}
suppressPackageStartupMessages(library(MASS))
suppressPackageStartupMessages(library(ggplot2))
suppressPackageStartupMessages(library(lme4))
suppressPackageStartupMessages(library(xtable))
suppressPackageStartupMessages(library(plyr))
suppressPackageStartupMessages(library(reshape))
suppressPackageStartupMessages(library(lattice))
suppressPackageStartupMessages(library(LMERConvenienceFunctions))
suppressPackageStartupMessages(library(car))
suppressPackageStartupMessages(library(grid))
suppressPackageStartupMessages(library(lmerTest))
suppressPackageStartupMessages(library(reshape2))
suppressPackageStartupMessages(library(ggplot2))
suppressPackageStartupMessages(require(languageR))
```

# Experiment Structure
Each subject saw 192 trials, being 3 cycles of 64 trials. There were 38 participants. There were 7296 rows in the full data set.

# Trial Procedure
The procedure for each trial was as follows. First an instruction that was a referring expression was presented until participants dismissed it with a key press. Then the squares with dots were presented until the participant responded with a key press, with a 60 second timeout if they did not respond.

# Prepare the data
Source additional R files with functions for computing confidence intervals.

```{r sourceFunctions}
source("functions/CIfunctions.R")
```

Get data.

```{r readdata}
dat <- read.delim("data/allresults.txt", stringsAsFactors = FALSE)

# declare local variables
number_of_valid_subjects <- 38
number_of_rows <- 7296
number_of_trials_per_subject <- number_of_rows / number_of_valid_subjects # 192

# sort out borderline responses into expected near and far
# How many squares in the square they chose? 
dat$RESPONSE <- as.character(dat$RESPONSE)
for (row in 1 : nrow(dat) ) {
  switch(dat[row,'RESPONSE'],
         'LEFT' = {dat[row, 'choice'] <- dat[row, 'Left']},
         'MIDDLE' = {dat[row, 'choice'] <- dat[row, 'Mid']},
         'RIGHT' = {dat[row, 'choice'] <- dat[row, 'Right']}         
  )
}
dat$crossed=paste('Con',dat$Condition,':Quan',dat$Quantity,':Item',dat$Item,sep="")

@@

## treat variables

# SUBJECT
# ensure Subject is a factor 
dat$Subject=factor(paste("s",sprintf("%02d",dat$Subject),sep=""))

# TRIAL
# Trial for subject, 1 to 192
dat$Trial = rep(x = 1:number_of_trials_per_subject, times = number_of_valid_subjects)

@@

# VAGUENESS
# Create a factor coding for Vagueness
dat[ dat$Condition==1 , 'Vagueness'] <- 'Vague'
dat[ dat$Condition==2 , 'Vagueness'] <- 'Crisp'
dat[ dat$Condition==3 , 'Vagueness'] <- 'Vague'
dat[ dat$Condition==4 , 'Vagueness'] <- 'Crisp'
dat$Vagueness <- as.factor(dat$Vagueness)

@@

dat$Selection = ""
dat[dat$Condition %in% c(1,2), "Selection"] <- "matching"
dat[dat$Condition %in% c(3,4), "Selection"] <- "comparison"
dat$Selection.sum <- ifelse(dat$Selection=="matching", .5, -.5)
dat$Selection <- as.factor(dat$Selection)

# remove one zero RT by replacing with NA
dat$RT[dat$RT==0]<-NA

# ORDER
# give the levels of Order meaningful names
dat$Order <- factor(dat$Order, levels=c(1,2), labels=c('LtoR','RtoL'))

@@

# QUANTITY
# give the levels of Quantity meaningful names
dat$Quantity <- factor(dat$Quantity, levels=c(1,2), labels=c('Small','Large'))

@@

# INSTRUCTION
# add number of characters in the instruction # 29 30 34 36 38
dat$nchar_instr = nchar(as.character(dat$Instruction))
dat$nchar_instr_scaled = as.vector(scale(nchar(as.character(dat$Instruction)), scale=TRUE))

# make Instruction be a factor (17 levels)
dat$Instruction <- as.factor(dat$Instruction)

# RT
# add transformations of RT
dat$RT_RecSqd    <- 1/dat$RT^2
dat$RT_Rec       <- 1/dat$RT
dat$RT_RecSqt    <- 1/sqrt(dat$RT) 
dat$RT_log       <- log(dat$RT)
dat$RT_sqt       <- sqrt(dat$RT)
dat$RT_raw       <- dat$RT
dat$RT_sqd       <- dat$RT^2

# save dat as dat.Rda
# This data set contains *all* trials including impossible trials and is mainly for graphs comparing different removals
save(dat, file="dat.Rda")

 # lose impossible trials
  dd <- dat
  dd$RT[dd$RT == 1 ] <- NA
  dd$RT[dd$RT > 40000 ] <- NA
  dd <- dd[complete.cases(dd),]
  row.names(dd) <- NULL

# add preceding RT: because we removed impossible trials, the value for preceding RT for a trial following an impossible trial is the value of the trial that preceded the impossible trial.
dd$RTprev <- NA
for (s in levels(dd$Subject)) {
  nrows = nrow(dd[dd$Subject==s,])
  for (i in 1:nrows) {
    if (i==1) {
      dd[dd$Subject==s, "RTprev"][i] <- dd[dd$Subject==s, "RT"][i]
    }
    else
      dd[dd$Subject==s, "RTprev"][i] <- dd[dd$Subject==s, "RT"][i-1]
  }
}

# add transformations of previous RT
dd$RTprev_RecSqd    <- 1/dd$RTprev^2
dd$RTprev_Rec       <- 1/dd$RTprev
dd$RTprev_RecSqt0   <- 1/sqrt(dd$RTprev) 
dd$RTprev_RecSqt    <- as.vector(scale( 1/sqrt(dd$RTprev)  ))
dd$RTprev_log       <- log(dd$RTprev)
dd$RTprev_sqt       <- sqrt(dd$RTprev)
dd$RTprev_raw       <- dd$RTprev
dd$RTprev_sqd       <- dd$RTprev^2

# save dd as dd.Rda
# This data set non-impossible trials and renamed/recomputed variables
save(dd, file="dd.Rda")
```

Check the distribution of the trimmed data to see if a transformation is suggested. Log transformation seems to be best so we adopt that for the rest of the analyses.

```{r boxplot subjects and items}
load("dd.Rda")
# allow reference to transformation
subdata = melt(dd,
               measure.vars=c("RT_Rec", "RT_RecSqt", "RT_log", "RT_raw"),
               variable.name="transformation",
               value.name="value") 
# boxplots for subjects and items separately for each dependent variable
ggplot(subdata) + 
  facet_wrap(~ variable, scales="free", ncol=4) +
  geom_boxplot(aes(y=value, x=Item, col="Items")) + 
  geom_boxplot(aes(y=value, x=Subject, col="Subjects")) + 
  theme_bw() +
  theme(axis.text.x = element_text(angle = 90, hjust = 1), aspect.ratio=1, panel.grid=element_blank()) + 
  scale_colour_manual("",values=c("Items"="red","Subjects"="blue")) +
  xlab("") + ylab("")
```

```{r load_dd}
load("dd.Rda")
```

```{r do_lmer}
v5 <- lmer(data=dd, 
           RT_RecSqt ~ 
             c_Vag + Selection.sum + c_Qty + c_Ord + 
             Selection.sum:c_Vag:c_Qty +
             item_mean_ratio + 
             s_Trl +
             RTprev_RecSqt +
             nchar_instr +
             (1+c_Vag + Selection.sum + c_Qty + c_Ord|Subject)
)
```

```{r summary_lmer}
summary(v5)
```

```{r R2}
cat("R^2")
cor(fitted(v5), dd$RT_RecSqt)^2
```

```{r plotLMER}
par(mfrow=c(2,5))
plotLMER.fnc(v5)
```

Plot model coefficients and ci's

```{r coeff}
par(las=2, mar=c(14,6,1,1))
y=as.data.frame(summary(v5)$coef[-1,1]) # estimates, without intercept
q=confint(v5)[-c(1,2,3),]
y=cbind(y,q)
plot(y[,1], ylim=range(y[,1],y[,2],y[,3]), xaxt="n", xlab="", ylab="estimate\n",  pch='-', cex=2)
segments(x0=1:nrow(y), x1=1:nrow(y), y[,2], y[,3])
abline(h=0)
axis(1, labels=row.names(y), at=1:nrow(y), srt=45)
grid()
```

Model criticism baayen plots

```{r baayenplots, fig.width=9, fig.height=3}
# Baayen 4-plot model criticism

par(mfrow=c(1,4), pty='s')
# create scaled residuals
v5$rstand = as.vector(scale(resid(v5)))
# plot scaled residuals density
plot(density(v5$rstand))
# plot sample quantiles versus theoretical quantiles
qqnorm(v5$rstand, cex=.5)
qqline(v5$rstand)
# plot standardised residuals versus fitted values
plot(v5$rstand ~ fitted(v5), pch='.')
# absolute standardised residuals greater than 2.5 are candidates for being outliers, the abline identifies them on the plot
abline(h=c(-2.5,2.5))
# create dffits
dffits=abs(resid(v5, "dffits"))
# plot dffits
plot(dffits, type='h')
```

>>>>

Get data from the raw per-subject files if necessary.

```{r, getData}
if (file.exists('rawdata.txt')) {
  dat <- read.delim(file="rawdata.txt", header=T, stringsAsFactors=FALSE)
} else {
  system(wait=TRUE, paste("head -n 1 ", "../../experimentCode","/output/", "subject01.data ", "> ", "rawdata.txt", sep=""))
  system(wait=TRUE, paste("tail -q -n +2 ", "../../experimentCode", "/output/", "*.data ",  ">> ", "rawdata.txt", sep=""))
  dat <- read.delim(file="rawdata.txt", header=T, stringsAsFactors=FALSE)
}

# declare local variables
Selection_of_valid_subjects <- 40
Selection_of_rows <- 7680
Selection_of_trials_per_subject <- Selection_of_rows / Selection_of_valid_subjects # 192

# sort out borderline responses into expected near and far
# How many squares in the square they chose? 
dat$RESPONSE <- as.character(dat$key)
for (row in 1 : nrow(dat) ) {
  switch(dat[row,'RESPONSE'],
         'L' = {dat[row, 'choice'] <- dat[row, 'Lft']},
         'M' = {dat[row, 'choice'] <- dat[row, 'Mid']},
         'R' = {dat[row, 'choice'] <- dat[row, 'Rgt']}         
  )
}
dat$Item <- dat$Itm
dat$Itm <- NULL
dat$crossed=paste('Con',dat$Cnd,':Quan',dat$Qty,':Item',dat$Item,sep="")

@@

# SUBJECT
# ensure Subject is a factor 
dat$Subject=factor(paste("s",sprintf("%02d",dat$Sub),sep=""))

# TRIAL
# Trial for subject, 1 to 192
dat$Trial = rep(x = 1:Selection_of_trials_per_subject, times = Selection_of_valid_subjects)

@@

# VAGUENESS
# Create a factor coding for Vagueness
dat[ dat$Cnd==1 , 'Vagueness'] <- 'Vague'
dat[ dat$Cnd==2 , 'Vagueness'] <- 'Crisp'
dat[ dat$Cnd==3 , 'Vagueness'] <- 'Vague'
dat[ dat$Cnd==4 , 'Vagueness'] <- 'Crisp'
dat$Vagueness <- as.factor(dat$Vagueness)

@@

dat$Selection = ""
dat[dat$Cnd %in% c(1,2), "Selection"] <- "Matching"
dat[dat$Cnd %in% c(3,4), "Selection"] <- "Comparison"
dat$c_Sel <- ifelse(dat$Selection=="Matching", .5, -.5)
dat$Selection <- as.factor(dat$Selection)

# remove redundant confusing columns
dat$Sub <- NULL
dat$Trl <- NULL

# ORDER
# give the levels of Order meaningful names
dat$Order <- factor(dat$Ord, levels=c(1,2), labels=c('LtoR','RtoL'))

@@

# QUANTITY
# give the levels of Quantity meaningful names
dat$Quantity <- factor(dat$Qty, levels=c(1,2), labels=c('Small','Large'))

@@

# INSTRUCTION
# add Selection of characters in the instruction # 29 30 34 36 38
dat$nchar_instr = nchar(as.character(dat$Ins))
dat$nchar_instr_scaled = as.vector(scale(nchar(as.character(dat$Ins)), scale=TRUE))

# make Instruction be a factor (17 levels)
dat$Ins <- as.factor(dat$Ins)

# RT
# add transformations of RT
dat$RT <- dat$rt; dat$rt <-  NULL
dat$RT_RecSqd    <- 1/dat$RT^2
dat$RT_Rec       <- 1/dat$RT
dat$RT_RecSqt    <- 1/sqrt(dat$RT) 
dat$RT_log       <- log(dat$RT)
dat$RT_sqt       <- sqrt(dat$RT)
dat$RT_raw       <- dat$RT
dat$RT_sqd       <- dat$RT^2

# lose impossible trials
dd <- dat
dd$RT[dd$RT > 9000 ] <- NA
dd <- dd[complete.cases(dd),]
row.names(dd) <- NULL

# # add preceding RT: because we removed impossible trials, the value for preceding RT for a trial following an impossible trial is the value of the trial that preceded the impossible trial.
# dd$RTprev <- NA
# for (s in levels(dd$Subject)) {
#   nrows = nrow(dd[dd$Subject==s,])
#   for (i in 1:nrows) {
#     if (i==1) {
#       dd[dd$Subject==s, "RTprev"][i] <- dd[dd$Subject==s, "RT"][i]
#     }
#     else
#       dd[dd$Subject==s, "RTprev"][i] <- dd[dd$Subject==s, "RT"][i-1]
#   }
# }
# 
# # add transformations of previous RT
# dd$RTprev_RecSqd    <- 1/dd$RTprev^2
# dd$RTprev_Rec       <- 1/dd$RTprev
# dd$RTprev_RecSqt0   <- 1/sqrt(dd$RTprev) 
# dd$RTprev_RecSqt    <- as.vector(scale( 1/sqrt(dd$RTprev)  ))
# dd$RTprev_log       <- log(dd$RTprev)
# dd$RTprev_sqt       <- sqrt(dd$RTprev)
# dd$RTprev_raw       <- dd$RTprev
# dd$RTprev_sqd       <- dd$RTprev^2

dd$Condition <- as.factor(paste(dd$Selection,dd$Vagueness, sep=' '))
dd$Target <- dd$Prm
foo <- subset(dd, select=c(
  "Subject", "Trial", "Condition", "Order", "Quantity", "Vagueness", "c_Vag", "Selection", "c_Sel", "Item", "c_Itm", "Target", "Ins", "switchResponse", "RT", "RT_log"
))
write.table(foo, file='../data.txt')
```
